---
title: "Fatigue"
output: html_document
date: "2024-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in Data
```{r}
savant_pitch_level <- read_csv("savant_pitch_level.csv")
fangraphs_season_level <- read_csv("fangraphs_season_level.csv")
```

# Select variables of savant of interest
```{r}
pitches <- savant_pitch_level %>% 
  select(
    pitcher, game_date, pitch_number_appearance, pitcher_at_bat_number, 
    pitch_type, release_speed, release_pos_x, release_pos_z,
    pfx_x, pfx_z, plate_x, plate_z, vx0, vy0, vz0, ax, ay, az, effective_speed, release_spin_rate,
    release_extension, release_pos_y
         ) %>%
  na.omit() %>%
  mutate(year = as.numeric(substr(game_date, 1, 4))) %>% select(-game_date)
```

# Response variables to calculate a fatigue rating for
```{r}
responses <- c("release_speed", "release_pos_x", "release_pos_z", "pfx_x", "pfx_z", "plate_x", "plate_z", "vx0", "vy0", "vz0", "ax", "ay", "az", "effective_speed", "release_spin_rate", "release_extension", "release_pos_y")
```

# Calculate fatigue ra†ing for each pitcher and pitch type for response variables
```{r}
pitcherids <- unique(pitches$pitcher)
fatigues_pitchers <- data.frame(pitcherid = integer(), 
                                pitch_type = character(), 
                                fatigue_coef = numeric())

total <- length(pitcherids)
pb <- txtProgressBar(min = 0, max = total, style = 3)


for (i in seq_along(pitcherids)) {
  pitcher_data <- pitches %>% filter(pitcher == pitcherids[i])
  pitcher_years <- unique(pitcher_data$year)
  pitcher_types <- unique(pitcher_data$pitch_type)
  # if pitcher_data empty skip
  if (nrow(pitcher_data) == 0) {
    setTxtProgressBar(pb, i)
    next
  }
  for (pitcher_year in pitcher_years) {
    pitcher_year_data <- pitcher_data %>% filter(year == pitcher_year)
    for (pitcher_type in pitcher_types) {
      for (response in responses) {
        pitcher_type_data <- pitcher_year_data %>% filter(pitch_type == pitcher_type)
        if (nrow(pitcher_type_data) == 0) {
          next
        }
        pitchermod <- lm(as.formula(paste(response, "~ pitch_number_appearance")), data = pitcher_type_data)
        fatigues_pitchers <- rbind(fatigues_pitchers, 
                                   data.frame(
                                     pitcherid = pitcherids[i], 
                                     year = pitcher_year, 
                                     pitch_type = pitcher_type,
                                     response_type = response, 
                                     fatigue_coef = coef(pitchermod)["pitch_number_appearance"])
                                   )
      }
    }
  }
  # Update the progress bar after each iteration
  setTxtProgressBar(pb, i)

  
}
close(pb)
```

# Pivot the data to wide format
```{r}
fatigues_full <- fatigues_pitchers %>% 
  filter(!pitch_type %in% c("PO", "FO", "KN", "SC", "EP", "UN", "FS", "SF", "KC")) %>%
  pivot_wider(names_from = pitch_type, values_from = fatigue_coef, names_prefix = "fatigue_") %>%
  pivot_wider(names_from = response_type, values_from = contains("fatigue_"))
```

# Join the fatigue data to the fangraphs data
```{r}
fangraphs_season_level <- left_join(fangraphs_season_level, fatigues_full, 
                               by = c("MLBAMID" = "pitcherid", "Season" = "year"))
```


# Write CSV of data before missing data imputation
```{r}
#write.csv(fangraphs_season_level, "preimputation_fangraphs.csv")
```




**Load in preimputation_fangraphs.csv if you want to skip above code**
```{r}
preimputation_fangraphs <- read_csv("preimputation_fangraphs.csv")
```

# Calculate Percentiles
```{r}
percentiles <- preimputation_fangraphs %>% filter(Season == 2023) %>%
  filter(GS >= 5) %>%
  filter(Role == "SP") %>%
  select(PlayerId, Name, Season, Age, contains("fatigue")) %>%
  mutate(across(contains("fatigue"), ~ percent_rank(.)*100)) 
```

# Calculate Fatigue rating by taking the mean of all percentiles
```{r}
fatigue_rating <- percentiles %>%
  pivot_longer(cols = contains("fatigue"), names_to = "variable", values_to = "values") %>%
  group_by(PlayerId, Name, Age) %>%
  summarise(fatigue_rating = mean(abs(values), na.rm = T))


# if you just want a rating for spin rate for example
#mean_spin_fatigue <- percentiles %>%
 # pivot_longer(cols = contains("release_spin_rate"), names_to = "variable", values_to = "values") %>%
#  group_by(PlayerId, Name) %>%
 # summarise(mean_velo_fatigue = mean(abs(values), na.rm = T))
```

# Read in full imputed data
```{r}
imputed_pitches <- read_csv("imputed_pitches.csv")
```

# Create a dataset of only SP information to predict fatigue_rating
```{r}
SP_mod_df <- left_join(fatigue_rating, imputed_pitches %>% select(-1)) %>% na.omit() %>%
  select(-c(GS, W, L, CG, ShO, SV, BS, HLD, SD, MD, Pulls, ERA)) %>%
  mutate(across(c(IP:Pitches, -TBF), ~./TBF, .names = "{.col}_TBF")) %>%
  mutate(across(c(G, H:Pitches), ~./IP, .names = "{.col}_IP")) %>%
  mutate(across(c(IP:Pitches), ~./G, .names = "{.col}_G")) %>%
  select(-Throws, -c(G:RS)) %>% ungroup() 
```

# Build a Lasso regression model predicting fatigue for SP

```{r}
X <- model.matrix(fatigue_rating ~ . -1, data = SP_mod_df %>% select(-PlayerId, -Name)) 
y <- SP_mod_df$fatigue_rating

set.seed(123) 
cv_lasso <- cv.glmnet(X, y, alpha = 0) 

best_lambda <- cv_lasso$lambda.min

lasso_best <- glmnet(X, y,  alpha = 0, lambda = best_lambda)

trainIndex <- createDataPartition(y, p = .75, list = FALSE)
X_train <- X[trainIndex, ]
y_train <- y[trainIndex]
X_test <- X[-trainIndex, ]
y_test <- y[-trainIndex]

# Refit the model on training data
cv_lasso_train <- cv.glmnet(X_train, y_train,  alpha =0)

# Predict on test data
predictions <- predict(cv_lasso_train, newx = X_test, s = "lambda.min", type = "response")

# Calculate the squared differences
squared_errors <- (predictions - y_test)^2

# Calculate the mean of these squared differences (MSE)
mse <- mean(squared_errors)

# Print the MSE
print(paste("MSE:", mse))

ggplot(data.frame(predictions = predictions, y_test = y_test), aes(x = predictions, y = y_test)) + geom_point() +
  geom_abline()
```


# Get dataset of relievers to predict fatigue rating
```{r}
RP_mod_df <- left_join(fangraphs_season_level %>% select(PlayerId, Name, Age, Season, Role), 
                       imputed_pitches %>% select(-1)) %>%
  na.omit() %>%
  filter(Season == 2023) %>%
  distinct(PlayerId, .keep_all = T) %>%
  filter(GS < 5) %>%
  filter(Age < 30) %>%
  select(-c(GS, W, L, CG, ShO, SV, BS, HLD, SD, MD, Pulls, ERA)) %>%
  mutate(across(c(IP:Pitches, -TBF), ~./TBF, .names = "{.col}_TBF")) %>%
  mutate(across(c(G, H:Pitches), ~./IP, .names = "{.col}_IP")) %>%
  mutate(across(c(IP:Pitches), ~./G, .names = "{.col}_G")) %>%
  select(-Throws, -c(G:RS), -Season, -Role) %>% ungroup() 
```

# Make predictions on the relievers expected fatigue rating using the starter model

```{r}
X_RP <- model.matrix(~ . -1, data = RP_mod_df %>% select(-PlayerId, -Name))

# Use the SP model to predict fatigue ratings for RPs
predictions_RP <- predict(lasso_best, newx = X_RP, s = "lambda.min", type = "response")

RP_predicted_fatigue <- data.frame(
  PlayerId = RP_mod_df$PlayerId,
  Name = RP_mod_df$Name,
  Age = RP_mod_df$Age,
  Expected_Fatigue_Rate = c(predictions_RP)
)
```

